UCD\\MSc Thesis Proposal:\\\emph{An approach for Continuous Capacity Planning in Cloud Environments with Uptime-based Pricing Model}

### 2015-04-23 paper/paper.tex
updates

\subsubsection{New capacity configuration suggestion}

@@

Evaluation process can be splitted into four phases:
\begin{enumerate}
\item collection and transformation of pricing and usage data,
\item calculation of key indicators (\textit{cph, ip(scheme1, scheme2), KDI}) and cluster characteristics (\textit{total cost, performance averages}), 
\item suggestion of changes based on Algorithm~\ref{alg:squeeze-them-in} and~\ref{alg:new_configuration},
\item validation of positive correlation between \textit{KDI} and \textit{total cost} changes.
\end{enumerate}

@@

\myparagraph{Platform}
Apache Spark 1.2.1\footnote{\url{https://spark.apache.org}} - an engine for large-scale data processing. All of map reduce jobs were written using Python API. Jobs were executed using standalone cluster mode with settings specified in Table~\ref{tab:spark_conf}. Job run times varied widely based on input data size and complexity of computations - from 20hs+ for transformations on 100G+ data to 1 minute aggregations on 100-200M input.

@@

For evaluation purposes we choose two pricing schemes for two different VM configurations~\cite{AWS:light}~\cite{AWS:medium}~\cite{AWS:heavy}. Pricing details are specified in Table~\ref{tab:aws_pricing}

@@

Google released 29-day cluster usage traces of more than 12k machines running over 650k jobs~\cite{clusterdata:Reiss2011}. An analysis~\cite{clusterdata:Reiss2012b} revealed that the workload is heterogeneous, which is somewhat unique compared to other traces and makes a good fit for testing.
The trace contains data about jobs submitted by users. Each job is made of tasks. Each task reports metrics about resource consumption - e.g. CPU, memory - and timestamps. The trace does not contain information about the purpose of the jobs, which limits SLO assumptions to resource utilization. 
Resource utilization data is provided in normalized units - the strongest configuration is 1 and the rest of configurations are described in relation to that value.

@@

Google trace data contains a lot of detailed information. This paper focuses on uptime and we extracted only relevant data for further evaluation. Machine and task events original schema is presented on Tables~\ref{tab:machine_events} and~\ref{tab:task_events} and fields selected for further transformation and aggregation are marked in bold text.

@@

\myparagraph{Uptime}
Uptime per day per machine was calculated based on aggregation of task duration times for a given day. Algorithm~\ref{alg:uptime_per_machine} shows design and Listing~\ref{uptime_per_machine_implementation} represents implementation of this aggregation. Total number of hours in cluster is 8705743, which gives 23.36 hours daily per machine on average.

@@

Based on Formula~\ref{eq:cph} implemented in Python and presented on Listing~\ref{cost_per_hour}, we calculated \textit{cost per hour} for both analyzed schemes. Results are presented on Table~\ref{tab:cph:scheme1} and~\ref{tab:cph:scheme2} in respect of number of hours per day a machine would run. The cost per hour drops when uptime per day rises. The same trend exhibits for both schemes.

@@

Based on Formula~\ref{alg:intersection points} implemented in Python and presented on Listing~\ref{intersection_points} we calculated \textit{intersection points} between pricing schemes. Table~\ref{tab:intersection_points} shows results of the calculation. Intuitively, heavy schemes is the cheapest option for highly utilized machines. An interesting finding is that medium is never cheaper than light in case of Scheme 2. Figures~\ref{fig:cc2_8xlarge} and~\ref{fig:m2_4xlarge} represent graphically \textit{cph} results and \textit{intersection points} can be noticed.

@@

Key distance indicator was calculated based on Algorithm~\ref{alg:distance_from_optimality}. 
Total cost was calculated based on total uptime and pricing scheme assignment. Results are presented on Table~\ref{tab:kdi_and_cost}.

@@

\begin{table}[h]
\begin{center}
    \begin{tabular}{| l | l | l |}
    \hline
    \textbf{Property} & \textbf{Scheme 1} & \textbf{Scheme 2} \\
    \hline
    Number of machines to tune & 932 & 937 \\
    \hline
    Total cost & \$657 200 & \$532 840.76 \\
    \hline
    CPU capacity to accommodate & 1 320 781.36 & 1 327 202.18 \\
    \hline
    Memory capacity to accommodate & 4 286 742.06 & 4 303 983.31 \\
    \hline
    Number of tasks to execute & 118869614 & 119367988 \\
    \hline
    Total uptime & 648519 & 651899 \\
    \hline
    Average cost per hour & \$1.0134 & \$0.817 \\
    \hline
    Average cost per task & \$0.00553  & \$0.00446 \\
    \hline
    \end{tabular}
\end{center}
\caption{Results of report run on cluster with criteria of KDI below 0 for 29 days}
\label{tab:threshold_based_report}
\end{table}

The cost of running such identified non-optimal machines is around 8\% and 10\% of total cost for Scheme 1 and 2 respectively.

@@

Once inefficiencies are identified, we will attempt to find a more optimal solution.

\myparagraph{Scheme 1}
total spare cpu = 3 697 171.44
total spare mem = 6 685 407.83
new KDI: +3 500 037
old KDI: +3 215 231
KDI change: +8.14\%
total uptime = 8056601
new total cost = \$8 177 048 - \$657 200 = \$7 519 848
cost change: -8.04\%
cost per hour = \$7 519 848 / 8056601 = 0.93337723935938
new cost per task = \$7 519 848 / 118869614 = 0.06326131419927

new averages 
cpu per machine = 15972217.38/11651 = 1370.88 / 29 = 47.27 +7.4\%
Avg mem usage per machine = 30676363.36/11651 = 2632.94/29 = 90.79 +7.41\%
Avg tasks number per machine = 1234716525/11651 = 105975.15/29 = +7.4\%

\myparagraph{Scheme 2}
total spare cpu = 3 694 240.56 
total spare mem = 6 682 956.59
new KDI: +5 913 933
old KDI: +5 431 137
KDI change: +8.16\%
total uptime = 8053221
new total cost = \$5 637 418 - \$532 840.76 = \$5 104 577.24
cost change: -9.45\%
cost per hour = \$5 104 577.24 / 8053221 = 0.63385535303204
new cost per task = \$5 104 577.24 / 119367988 = 0.04276336834964

new averages 
cpu per machine = 15972217.38/11646 = 1371.48 / 29 = 47.29 +7.44\%
Avg mem usage per machine = 30676363.36/11646 = 2634.07 / 29 = 90.83 +7.45\%
Avg tasks number per machine = 1234716525/11646 = 106020.65 / 29 = 3655.88 +7.45\%

@@

$vm1 = Entity(c_name, c_light, c_medium, c_heavy)$
$vm2 = Entity(m_name, m_light, m_medium, m_heavy)$

$calc_options(uptime, vm2, 29)$
(1597, 3727, 972, 567 273.49, 998 186.23, 410 041.68)
$calc_options(uptime, vm1, 29)$
(1397, 1720, 972, 770 417.58, 876 376.34, 604 274.61)

\section{Validate influence of suggestions and correlation with KDI}

>>>>

\subsubsection{New Capacity configuration suggestion}

@@

Evaluation process can be splitted into four phases:
\begin{enumerate}
\item collection and transformation of pricing and usage data,
\item calculation of key indicators (\textit{cph, ip(scheme1, scheme2), KDI}) and cluster characteristics (\textit{total cost, performance averages}), 
\item suggestion of changes based on Algorithm~\ref{alg:squeeze-them-in} and~\ref{alg:new_configuration},
\item validation of a positive correlation between \textit{KDI} and \textit{total cost} changes.
\end{enumerate}

@@

\myparagraph{Platform}
Apache Spark 1.2.1\footnote{\url{https://spark.apache.org}} - an engine for large-scale data processing. All of the map-reduce jobs were written using Python API. Jobs were executed using standalone cluster mode with settings specified in Table~\ref{tab:spark_conf}. Job run times varied widely based on input data size and complexity of computations - from 20hs+ for transformations on 100G+ data to 1-minute aggregations on 100-200M input.

@@

For evaluation purposes, we choose two pricing schemes for two different VM configurations~\cite{AWS:light}~\cite{AWS:medium}~\cite{AWS:heavy}. Pricing details are specified in Table~\ref{tab:aws_pricing}

@@

Google released 29-day cluster usage traces of more than 12k machines that run over 650k jobs~\cite{clusterdata:Reiss2011}. An analysis~\cite{clusterdata:Reiss2012b} revealed that the workload is heterogeneous, which is somewhat unique compared to other traces and makes a good fit for testing.
The trace contains data about jobs submitted by users. Each job is made of tasks. Each task reports metrics about resource consumption - e.g. CPU, memory - and timestamps. The trace does not contain information about the purpose of the jobs, which limits SLO assumptions to resource utilization. 
Resource utilization data is provided in normalized units - the strongest configuration is 1 and the rest of configurations are described in relation to that value.

@@

Google trace data contains a lot of detailed information. This paper focuses on uptime and we extracted only relevant data for further evaluation. Machine and task events original schema is presented in Tables~\ref{tab:machine_events} and~\ref{tab:task_events} and fields selected for further transformation and aggregation are marked in bold text.

@@

\myparagraph{Uptime}
Uptime per day per machine was calculated based on an aggregation of task duration times for a given day. Algorithm~\ref{alg:uptime_per_machine} shows design and Listing~\ref{uptime_per_machine_implementation} represents the implementation of this aggregation. The total number of hours in the cluster is 8705743, which gives 23.36 hours daily per machine on average.

@@

Based on Formula~\ref{eq:cph} implemented in Python and presented in Listing~\ref{cost_per_hour}, we calculated a \textit{cost per hour} for both analyzed schemes. Results are presented in Table~\ref{tab:cph:scheme1} and~\ref{tab:cph:scheme2} in respect of number of hours per day a machine would run. The cost per hour drops when uptime per day rises. The same trend exhibits for both schemes.

@@

Based on Formula~\ref{alg:intersection points} implemented in Python and presented in Listing~\ref{intersection_points} we calculated \textit{intersection points} between pricing schemes. Table~\ref{tab:intersection_points} shows results of the calculation. Intuitively, a heavy scheme is the cheapest option for highly utilized machines. An interesting finding is that a medium is never cheaper than light in case of Scheme 2. Figures~\ref{fig:cc2_8xlarge} and~\ref{fig:m2_4xlarge} represent graphically \textit{cph} results and \textit{intersection points} can be noticed.

@@

The key distance indicator was calculated based on Algorithm~\ref{alg:distance_from_optimality}. 
Total cost was calculated based on total uptime and pricing scheme assignment. Results are presented in Table~\ref{tab:kdi_and_cost}.

@@

\begin{table}[h]
\begin{center}
    \begin{tabular}{| l | l | l |}
    \hline
    \textbf{Property} & \textbf{Scheme 1} & \textbf{Scheme 2} \\
    \hline
    Number of machines to tune & 932 & 937 \\
    \hline
    Total cost & \$657 200 & \$532 841 \\
    \hline
    CPU capacity to accommodate & 1 320 781.36 & 1 327 202.18 \\
    \hline
    Memory capacity to accommodate & 4 286 742.06 & 4 303 983.31 \\
    \hline
    Number of tasks to execute & 118869614 & 119367988 \\
    \hline
    Total uptime & 648519 & 651899 \\
    \hline
    Average cost per hour & \$1.0134 & \$0.817 \\
    \hline
    Average cost per task & \$0.00553  & \$0.00446 \\
    \hline
    \end{tabular}
\end{center}
\caption{Results of report run on cluster with criteria of KDI below 0 for 29 days}
\label{tab:threshold_based_report}
\end{table}

The cost of running identified non-optimal machines is around 8\% and 10\% of total cost for Scheme 1 and 2 respectively.

@@

Based on Algorithm~\ref{alg:squeeze-them-in} we calculated spare capacity for both schemes. Results are presented in Table~\ref{tab:squeeze-them-in-capacity}

\begin{table}[h]
\begin{center}
    \begin{tabular}{| l | l | l |}
    \hline
    \textbf{Property} & \textbf{Scheme 1} & \textbf{Scheme 2} \\
    \hline
    Total spare CPU & 3 697 171.44 & 6 3 694 240.56  \\
    \hline
    Total spare memory & 6 685 407.83 & 6 682 956.59 \\
    \hline
    \end{tabular}
\end{center}
\caption{Results of spare capacity calculation on the cluster}
\label{tab:squeeze-them-in-capacity}
\end{table}

In both cases, there is more spare capacity than results from Table~\ref{tab:threshold_based_report} require. It indicates we can power off all of the non-optimal machines and investigate an impact on workload and pricing indicators. Results of such an experiment are presented in Table~\ref{tab:squeeze-them-in}.

\begin{table}[h]
\begin{center}
    \begin{tabular}{| l | l | l |}
    \hline
    \textbf{Property} & \textbf{Scheme 1} & \textbf{Scheme 2} \\
    \hline
    New KDI & +3 500 037 & +5 913 933 \\
    \hline
    KDI change & +8.14\% & +8.16\% \\
    \hline
    New total cost & \$7 519 848 & \$5 104 577.24 \\
    \hline
    Total cost change & -8.04\% & -9.45\% \\
    \hline
    New cost per hour & \$0.9334 & \$0.6339 \\
    \hline
    New cost per task & \$0.0633  & \$0.0428 \\
    \hline
    New CPU average per machine & 47.27 & 47.29 \\
    \hline
    CPU average change & +7.4\% & +7.44\% \\
    \hline
    New memory usage average & 90.79 & 90.83 \\
    \hline
    Memory average change & +7.41\% & +7.45\% \\
    \hline
    New task number average & 3654 & 3655 \\
    \hline
    Task average change & +7.4\% & +7.45\% \\
    \hline
    \end{tabular}
\end{center}
\caption{Cluster statistics with non-optimal machines powered off}
\label{tab:squeeze-them-in}
\end{table}

The KDI increased by above 8\% for both schemes. This is expected as we removed machines with negative KDI values. The total cost decreased by 8.04\% and 9.45\% for Scheme 1 and 2 respectively. This is also expected. We noticed a correlation between the increase of KDI and decrease of total cost. However, there is a trade-off in terms of resource utilization, which increased around 7.4\% on average. A user has to decide whether this is acceptable for his specific workload SLOs. It might work well for under-utilized clusters with a low peak-to-mean ratio. For high peak-to-mean workloads, the real difference in utilization patterns can be higher than detected by averages. To measure SLOs in these environments, we would have to use more specific metrics like requests per second or queries per second during peak.

@@

We used Algorithm~\ref{alg:new_configuration} to come up with new configuration suggestions based on the uptime of machines from Table~\ref{tab:threshold_based_report} and Scheme 1 and 2 pricing details. \\
The algorithm analyzed all three \textit{intersection points} available for each scheme and returned total cost and number of machines required. Results are presented in Table~\ref{tab:new-cap-scheme1} and~\ref{tab:new-cap-scheme2}.

\begin{table}[h]
\begin{center}
    \begin{tabular}{| l | l | l |}
    \hline
    \textbf{Scheme} & \textbf{Machines} & \textbf{Total cost} \\
    \hline
    Light & 1397 & \$770 417.58 \\
    \hline
    Medium & 1720 & \$876 376.34 \\
    \hline
    Heavy & 972 & \$604 274.61\\
    \hline
    \end{tabular}
\end{center}
\caption{New capacity configuration suggestions for Scheme 1}
\label{tab:new-cap-scheme1}
\end{table}

\begin{table}[h]
\begin{center}
    \begin{tabular}{| l | l | l |}
    \hline
    \textbf{Scheme} & \textbf{Machines} & \textbf{Total cost} \\
    \hline
    Light & 1597 & \$567 273.49 \\
    \hline
    Medium & 3727 & \$998 186.23 \\
    \hline
    Heavy & 972 & \$410 041.68 \\
    \hline
    \end{tabular}
\end{center}
\caption{New capacity configuration suggestions for Scheme 2}
\label{tab:new-cap-scheme2}
\end{table}

An interesting pattern raised - a medium is always the least optimal choice and heavy is the best. The suggestion to run heavy configuration for Scheme 1 would save 8.05\% compared to non-optimal machines set and 0.0065\% from total cost. Scheme 2 would save 23\% compared to non-optimal machines set and 0.022\% from total cost. KDI increased for both schemes. Scheme 1 noted 8.42\% change and Scheme 2 8.31\%. Light and medium suggestions turned out to be more expensive and were excluded from the rest of considerations. \\

As opposed to the previous algorithm, this one does not violate SLOs. The new suggested capacity handles the same amount of uptime using the same resource configuration as the rest of the cluster.

\section{Validate correlation of total cost and KDI}

\begin{figure}[H]
	\includegraphics[width=\linewidth]{figures/kdi_to_cost}
	\caption{KDI to total cost correlation}
	\label{fig:kdi_to_cost}
\end{figure}
